A Quantitative Genetic Linked Selection Model of the Negative Selection Process in Humans

### 2023-07-12 manuscript/manuscript.tex
ms updates

\begin{figure}[htbp] \centering
    \includegraphics[width=\textwidth]{figures/figure_2.pdf} 
    \caption{Comparison of parameter estimates using classic BGS theory (green
        lines) with our new B' method (blue lines) across both full and sparse
        track types (dark versus light hue), and different mutation rates
        (columns). In the first row, $R^2$ between predictions and observations
        increases with selection intensity. Both classic BGS and B' methods fit
        the data equally well with sparse tracks, though the estimated
        selection coefficients are inaccurate when selection is weak (second
        row). Mutation rate estimates (third row) are more accurately estimated
        by the B' method than classic BGS, but overall show slight biases.
        Overall, classic BGS methods fail as expected when full-coverage tracks
    are used, since it cannot accommodate weak selection and neutrality. }
  \label{fig:figure-2}
\end{figure}

@@

Overall, we find the PhastCons Priority and CADD 6\% models fit about equally
as well (Figure \ref{fig:figure-2}A), consistent with recent work using classic
BGS theory \parencite{Murphy2022-sj}. However, we find that our models predict
out-sample diversity levels slightly better than previous methods. For these
two models, we find that our B' method predicts $R_\text{LOCO}^2=67.3$\% and
$R_\text{LOCO}^2=68.7$\% of the out-sample variance in Yoruba pairwise
diversity at the megabase-scale, respectively. By contrast, the best-fitting
CADD 6\% model from \textcite{Murphy2022-sj} explained 60\% of diversity in
left-out 2 Mbp windows across YRI samples. We note that this difference could
be explained by other differences in data processing, optimization, etc.

In populations impacted by the out-of-Africa bottleneck, the goodness-of-fit
was lower across all models (e.g. 60.4\% and 58.7\% for CEU and CHB
respectively in the CADD 6\% model).

@@

Since our method is built upon theory that fixes the weak selection problem of
classic BGS theory, it should in principle fit equally well when an annotation
model includes regions that are under no selective constraint and thus
neutrally evolving. To test this, for each of our annotation models (which are
``sparse") we fit a corresponding ``full" model that assigns the remaining
complement of the genome to a feature called ``other" that contains most of the
genome. Ideally, how a model fits using our B' method should be invariant to
whether an annotation model is sparse or full, since our method in principle
can accommodate weak selection and neutrality. Indeed, we find this to be the
case: both in-sample $R^2$ and out-sample $R_\text{LOCO}^2$ values are nearly
identical across full and sparse-track models (Figures \ref{fig:figure-2}A and
\ref{fig:figure-2}B, round points).

By contrast, full annotation models fit poorly under classic BGS theory (Figure
\ref{fig:figure-2}B, diamond-shaped points) and lead to unreasonable parameter
estimates (discussed in Section XXX). Additionally, when sparse annotation
models contain genomic features that are likely under weak constraint (such as
introns and UTRs), models fit worse under classic BGS theory than our B' method
(Figure \ref{fig:figure-2}A). However, among the CADD annotation models, the
goodness-of-fit is nearly identical between B' and classic BGS methods. This
behavior is what we would expect given that the CADD models contain only the
most pathogenic sites, which are \emph{a priori} very likely under the strong
selection domain under which B' and classic BGS theory agree.

Overall, our $R_\text{LOCO}^2$ estimates suggest our negative selection model
explains up to 68\% of out-sample variance in diversity of the megabase scale,
even though our method assumes constant demography and homogeneous mutation
rates along the genome. A worthwhile question is: how much variation
\emph{could} we expect to fit at this scale? Given that selection alters
genealogies in ways beyond just decreasing mean pairwise coalescence time and
populations have non-constant demography, an exact analytic answer is
intractable. However, we can get an approximate idea if we assume that the
residual variance $\sum_b (\pi_b - \widehat{\pi}_b)^2$ is determined entirely
by the expected neutral coalescence noise around the expected coalescence time
$2B(b)N$. This can be be found analytically, plugging-in our predictions for
$B(b)$. This allows us to calculate $R_\text{coal}^2$ to ballpark the theoretic
variance that is capable of being explained, assuming this coalescence-only
noise process (see Method XXX). We note that selection is expected to
\emph{decrease} the variance in coalescence times beyond a rescaled effective
population size implies; our $R_\text{coal}^2$ would be an underestimate under
selection.

We find that our out-sample $R_\text{LOCO}^2$ for the Yoruba samples
($R_\text{LOCO}^2 \approx 68$\%) is slightly above the theoretic
$R_\text{coal}^2 \approx 67$\%. This suggests our model is close to the
vicinity of fitting all the signal possible, under the coalescence-only noise
assumption. By contrast, for bottlenecked out-of-Africa populations, we find a
larger discrepancy between $R_\text{coal}^2$ and observed out-sample
$R_\text{LOCO}^2$. The theoretic $R_\text{coal}^2 \approx 64$\% for both
populations, compared to the observed $R_\text{LOCO}^2 \approx 62$\% for CEU
and $R_\text{LOCO}^2 \approx 59$\% for CHB. Given that bottlenecks would act to
increase the residual variance in coalescence times beyond the level implied by
the effective population size, the gap would likely shrink under more realistic
models or simulation-based approximations for $R_\text{coal}^2$. Overall, this
suggests that negative selection models explain the vast majority coalescence
time variation at the megabase-scale that is capable of being explained (i.e.
that is not coalescence noise).

@@

Our composite likelihood method has three sets of parameters: expected
diversity in the absence of linked selection diversity $\pi_0$, the mutation
rate $\mu$, and the matrix of distribution of fitness effects $\mathbf{W}$
across the selection grid for each of the $K$ feature types. Given that the
relationship between $\pi$ and the strength selection is U-shaped, we wondered
whether the linked selection signal in humans could fit under a weak selection
regime.

However, across all of our annotation models, new deleterious mutations in
conserved feature classes were consistently estimated to have strongly
deleterious effects (Figure \ref{fig:figure-3}A), consistent with previous work
\parencite{McVicker2009-ax,Murphy2022-sj}. This includes features such as
coding sequences, PhastCons regions, or the top 6\% of basepairs with the most
pathogenic CADD scores.

@@

\begin{figure}[htbp] 
    \centering
    \includegraphics[width=\textwidth]{figures/figure_5.pdf} 
    \caption{(A) }
  \label{fig:figure-4}
\end{figure}

>>>>

\begin{figure}[htbp] \centering
    \includegraphics[width=0.8\textwidth]{figures/figure_2.pdf} 
    \caption{Comparison of parameter estimates using classic BGS theory (green
        lines) with our new B' method (blue lines) across both full and sparse
        track types (dark versus light hue), and different mutation rates
        (columns). In the first row, $R^2$ between predictions and observations
        increases with selection intensity. Both classic BGS and B' methods fit
        the data equally well with sparse tracks, though the estimated
        selection coefficients are inaccurate when selection is weak (second
        row). Mutation rate estimates (third row) are more accurately estimated
        by the B' method than classic BGS, but overall show slight biases.
        Overall, classic BGS methods fail as expected when full-coverage tracks
    are used, since it cannot accommodate weak selection and neutrality. }
  \label{fig:figure-2}
\end{figure}

@@

Overall, we find the PhastCons Priority and CADD 6\% models fit about equally
as well (Figure \ref{fig:figure-2}A), consistent with recent work using classic
BGS theory \parencite{Murphy2022-sj}. However, we find that our models predict
out-sample diversity levels slightly better than previous methods. For these
two models, we find that our B' method predicts $R_\text{LOCO}^2=67.3$\% and
$R_\text{LOCO}^2=66.7$\% of the out-sample variance in Yoruba pairwise
diversity at the megabase-scale, respectively. By contrast, the best-fitting
CADD 6\% model from \textcite{Murphy2022-sj} explained 60\% of diversity in
left-out 2 Mbp windows across YRI samples. We note that this difference could
be explained by other differences in data processing, optimization, etc. For
populations impacted by the out-of-Africa bottleneck, the goodness-of-fit was
lower across all models (e.g. 61.0\% and 58.8\% for CEU and CHB respectively in
the PhastCons Priority model).

@@

Since our method is built upon theory that fixes the weak selection problem of
classic BGS theory, it should in principle fit equally well when an annotation
model includes regions that are under no selective constraint and thus
neutrally evolving. To test this, for each of our annotation models (which are
``sparse") we fit a corresponding ``full" model that assigns the remaining
portion of the genome to a feature called ``other". Ideally, how a model fits
using our B' method should be invariant to whether an annotation model is
sparse or full, since our method in principle can accommodate weak selection
and neutrality. Indeed, we find this to be the case: both in-sample $R^2$ and
out-sample $R_\text{LOCO}^2$ values are nearly identical across full and
sparse-track models (Figures \ref{fig:figure-2}A and \ref{fig:figure-2}B, round
points).

By contrast, full annotation models fit poorly under classic BGS theory, and
lead to unreasonable parameter estimates (discussed in Section XXX).
Additionally, when sparse annotation models contain genomic features that are
likely under weak constraint (such as introns and UTRs), models fit worse under
classic BGS theory than our B' method (Figure \ref{fig:figure-2}A). However,
among the CADD annotation models, the goodness-of-fit is nearly identical
between B' and classic BGS methods. This behavior is what we would expect given
that the CADD models contain only the most pathogenic sites, which are \emph{a
priori} very likely under the strong selection domain under which B' and
classic BGS theory agree.

Overall, our $R_\text{LOCO}^2$ estimates suggest our negative selection model
explains up to 67\% of out-sample variance in diversity of the megabase scale,
even though our method assumes constant demography and homogeneous mutation
rates along the genome. A worthwhile question is: how much variation
\emph{could} we expect to fit at this scale? Given that selection alters
genealogies in ways beyond just decreasing mean pairwise coalescence time and
populations have non-constant demography, an exact analytic answer is
intractable. However, we can get an approximate idea if we assume that the
residual variance $\sum_b (\pi_b - \widehat{\pi}_b)^2$ is determined entirely
by the expected neutral coalescence noise around the expected coalescence time
$2B(b)N$. This can be be found analytically, plugging-in our predictions for
$B(b)$. This allows us to calculate $R_\text{coal}^2$ to ballpark the theoretic
variance that is capable of being explained, assuming this coalescence-only
noise process (see Method XXX). We note that selection is expected to
\emph{decrease} the variance in coalescence times beyond a rescaled effective
population size implies; our $R_\text{coal}^2$ would be an underestimate under
selection.

We find that our out-sample $R_\text{LOCO}^2$ for the Yoruba samples
($R_\text{LOCO}^2 \approx 67$\%) is slightly above the theoretic
$R_\text{coal}^2 \approx 66.6$\%. This suggests our model is in the vicinity of
fitting all the signal possible, under the coalescence-only noise assumption.
By contrast, for bottlenecked out-of-Africa populations, we find a larger
discrepancy between $R_\text{coal}^2$ and observed out-sample
$R_\text{LOCO}^2$. The theoretic $R_\text{coal}^2 \approx 63\%-64$\% for both
populations, compared to the observed $R_\text{LOCO}^2 \approx 61$\% for CEU
and $R_\text{LOCO}^2 \approx 59$\% for CHB under the PhastCons Priority model.
Given that bottlenecks would act to increase the residual variance in
coalescence times beyond the level implied by the effective population size,
this gap would likely shrink under more realistic models or simulation-based
approximations for $R_\text{coal}^2$. Overall, this suggests that negative
selection models explain the vast majority coalescence time variation at the
megabase-scale that is capable of being explained (i.e. that is not coalescence
noise).

@@

Our composite likelihood method has three sets of parameters: the expected
diversity in the absence of linked selection diversity $\pi_0$, the mutation
rate $\mu$, and the matrix of distribution of fitness effects $\mathbf{W}$
across the selection grid for each of the $K$ feature types. Given that the
relationship between $\pi$ and the strength selection is U-shaped, we wondered
whether our new B' model accommodating weak selection could fit the linked
selection signal under a different combination of weak and strong selection
parameters than found by previous work.

However, across all of our annotation models, new deleterious mutations in
conserved feature classes (CADD tracks, CDS, and PhastCons regions) were
consistently estimated to have strongly deleterious effects (Figure
\ref{fig:figure-4}A), consistent with previous work
\parencite{McVicker2009-ax,Murphy2022-sj}. The DFE estimates for CADD and
PhastCons regions consistently places $\ge 75\%$ of mass on the largest
selection coefficient, $s=10^{-2}$. The CADD 6\% DFE estimates imply an average
selection coefficient of $\bar{s} = 0.65\%$ for CEU, $\bar{s} = 0.57\%$ for
CHB, and $\bar{s} = 0.79\%$ for YRI. Similarly, the PhastCons Priority model
implies average selection coefficient estimates of $\bar{s} = 0.63\%$, $\bar{s}
= 0.58\%$, and $\bar{s}= 0.77\%$ for CEU, CHB, and YRI respectively for
PhastCons regions.

Following previous work, our method used a grid of selection coefficients up to
$s=10^{-2}$. However, we also experimented with a strong selection grid that
includes $s=10^{-1}$. We find that models fit with the strong selection grid
had $R_\text{LOCO}^2$ that were about one percentage point higher, which is
suggestive of stronger selection than has previously been estimated using
constrained grids (see Supplementary Materials Table
\ref{supp:tbl-r2-strongsel}). For this strong selection grid, we estimate
average selection coefficients for the CADD 6\% model of $\bar{s} = 4.3\%$ for
CEU, $3.2\%$ for CHB, and $4.4\%$ for YRI. Thus, across different selection
coefficient grids, there is considerable uncertainty in the average strength of
selection against deleterious mutations in conserved regions.

However, across the strong selection grid runs,

@@

\begin{figure}[htbp] 
    \centering
    \includegraphics[width=0.5\textwidth]{figures/figure_5.pdf}

    \caption{Mutation rate estimates across sparse track (top row) and
        full-coverage track (bottom row) models, for the new B' (circles) and
        classic BGS (diamonds) methods. Estimates of the mutation rate are
        consistent between classic BGS and B' methods for sparse tracks CADD
        models (overlapping diamonds and circles, top row). Overall, mutation
        rate estimates are sensitive to the underlying annotation model.}
  \label{fig:figure-4}
\end{figure}
